A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-13 paper/paper.tex
hide the acks for dbl blind
shorten the discussion section

The coend calculus~\citet{coend-calculus} is an informal syntax for
manipulating profunctors involving ends and coends; an extension of
\vett{} to treat profunctors of many variables of different variances
should would provide a formal treatment of it.

@@

Our work in this paper fits broadly into a line of work on
\emph{directed dependent type theories}, a type theory where the
identity type is interpreted as morphisms in a (possibly
$\infty$-)category.  In directed type theories based on a bisimplicial
model~\cite{riehl-shulman,weinberger,otherweinbegerpaper,licataweaver},
morphism types are defined using an interval object, like in cubical
type theory~\cite{bch,cchm,afh,abcfhl}, and universal properties like
``morphism induction'' are an internally definable property of certain
types.  Other type theories~\cite{north,ahrensnorth} define morphism
types via an induction principle, corresponding to the lifting
properties of certain kinds of fibrations of categories.  While these
previous works can express some constructions on $\Cat$ that are not
expressible in \vett{}, because \vett{} is more restricted, it has more
models, e.g. in internal and enriched categories, so the theorems that
are provable in \vett{} apply in more settings.

@@

\subsection{Category Theoretic Extensions}

While we have demonstrated that some common constructions in category
theory can be expressed using the simple combination of homs, tensors
and units, there are several that are out of reach for SCT due to its
restrictive judgmental structure. These require extensions to the
judgmental structure that would move beyond virtual equipments.

First, while we can add products to the calculus, adding in a
\emph{functor category} ${\cat C}^{\cat D}$ would necessitate adding a
syntax for functors of multiple arguments, to allow for the lambda rule:
\[ \inferrule {\alpha:\cat C, \beta:\cat D \vdash e : \cat E} {\alpha:\cat C \vdash \lambda \beta:\cat D. e : \cat E} \]
This would then require also some extension to profunctors for
allowing 0 or more covariant/contravariant dependencies as well.
%
Semantically, this would correspond to generalizing from virtual
equipments to some kind of (possibly cartesian) \emph{monoidal closed}
virtual equipments.

In addition, we could consider extending our semantics with a notion
of \emph{opposite category}, with the property that a covariant
dependence of a profunctor on a variable in a category should be
equivalent to a contravariant dependence on a variable in the opposite
category.
%
This could possibly be done by introducing a modal distinction
$\alpha:^{+}\cat C$ for covariant dependence and $\alpha:^{-}\cat C$
for contravariant dependence, with the opposite category being a
modality that internalizes flipping the variance as a type
constructor.
%
With these two extensions, possibly, the category contexts for
functors and profunctors could be unified as one notion of a sequence
of object variables with variances.
%
With this unification, it would then be simpler to incorporate a third
extension: a universe object, corresponding to an object
$\textrm{Set}$, which would allow us to internalize the profunctor
judgment in the functor judgment, modeling that in classical category
theory, profunctors are the same as certain functors to the category
of Sets.
%
While all of these extensions seem plausible, they all move away to
the simple ordered type theory structure that allowed us to define
units, tensor and hom profunctors so easily. It is unclear how to
define these types in this extended structure. As a simple example, if
profunctors have many variables of mixed variance, rather than exactly
one of each variance, it is unclear what the appropriate
generalization of the transformation context $\Phi$ should be.

An orthogonal extension to these would be to consider adding a
\emph{discretization} modality $\Delta\cat C$ that gives the discrete
category on the underlying set of objects of $\cat C$. This would give
us the ability to formalize concepts like ``function on objects'' or
``unnatural transformation'' without the need to define internal
categories. This could be modeled by adding a modal distinction
$\alpha:^{\#}\cat C$ for ``crisp'' dependence on an object of $\cat
C$, i.e., dependence merely on the object of $\cat C$ with no
functoriality/naturality conditions. This would then be internalized
by the discretization modality.

A third approach would be to extend the calculus to model general
monoidal closed structures and more generally multicategories, which
provides semantics of modal substructural type theories.
%
This can be modeled by extending the ``universe'' $\Set$ to be a
monoidal closed category, which induces a monoidal closed structure on
profunctors by the Day convolution product \cite{...}.
%
Similar developments along these lines have already been done for
coend calculus \citep{coend-calculus} and in formalization of
substructural type theories \citep{atkey-woods}.

@@

\subsection{Semantics}

Our definition of a double category with displays is suggestive of a
simple generalization. A double category is a monoid object in the
double category of $Span(Cat)$, and a double category with displays
furthermore uses certain diagrams involving profunctors, which also
form a double category $Prof$. This suggests that the algebraic
structure of a double category with displays is naturally modeled in
some kind of \emph{triple category}, modeling the three types of
morphisms: functors, profunctors and spans of functors.

@@

%% %% Please use bibtex, 
\subsubsection*{Acknowledgements}
Grants stuff...? The authors would like to thank David Jaz Myers, Emily Riehl, Mike
Shulman, Dominic Verity for helpful feedback on this work.

>>>>

The coend calculus~\citet{coend-calculus} is an informal syntax for
manipulating profunctors involving ends and coends; an extension of
\vett{} to treat profunctors of many variables of different variances
may provide a formal treatment of it.

@@

Our work in this paper fits broadly into a line of work on
\emph{directed dependent type theories}, a type theory where the
identity type is interpreted as morphisms in a (possibly
$\infty$-)category.  In directed type theories based on a bisimplicial
model~\cite{riehl-shulman,weinberger,otherweinbegerpaper,licataweaver},
morphism types are defined using an interval object, like in cubical
type theory~\cite{bch,cchm,afh,abcfhl}, and universal properties like
``morphism induction'' are an internally definable property of certain
types.  Other type theories~\cite{north,ahrensnorth} define morphism
types via an induction principle, corresponding to the lifting
properties of certain kinds of fibrations of categories.  While these
previous works can express some constructions on $\Cat$ that are not
expressible in \vett{}, because \vett{} is more restricted, \vett{}
contrariwise has more models, for instance categories enriched in
non-cartesian monoidal categories, so the theorems that are provable
in \vett{} apply in more settings.

@@

\subsection{Extending VETT}

While many constructions in category theory can be formalized in
virtual equipments, and therefore \vett{}, others require more
structure. Future extensions to \vett{} could incorporate functor
categories by generalizing the unary type theory of functors to
functors of many variables, in which case ordinary $\lambda$ calculus
can be used to define functor categories as function types. Similarly,
ideas from modal type theory/graded linear logic could be used for
defining opposite categories
\cite{mike-contravariance-enrichment,semiring-stuf??}.

%% While we have demonstrated that some common constructions in category
%% theory can be expressed using the simple combination of homs, tensors
%% and units, there are several that are out of reach for SCT due to its
%% restrictive judgmental structure. These require extensions to the
%% judgmental structure that would move beyond virtual equipments.

%% First, while we can add products to the calculus, adding in a
%% \emph{functor category} ${\cat C}^{\cat D}$ would necessitate adding a
%% syntax for functors of multiple arguments, to allow for the lambda rule:
%% \[ \inferrule {\alpha:\cat C, \beta:\cat D \vdash e : \cat E} {\alpha:\cat C \vdash \lambda \beta:\cat D. e : \cat E} \]
%% This would then require also some extension to profunctors for
%% allowing 0 or more covariant/contravariant dependencies as well.
%% %
%% Semantically, this would correspond to generalizing from virtual
%% equipments to some kind of (possibly cartesian) \emph{monoidal closed}
%% virtual equipments.

%% In addition, we could consider extending our semantics with a notion
%% of \emph{opposite category}, with the property that a covariant
%% dependence of a profunctor on a variable in a category should be
%% equivalent to a contravariant dependence on a variable in the opposite
%% category.
%% %
%% This could possibly be done by introducing a modal distinction
%% $\alpha:^{+}\cat C$ for covariant dependence and $\alpha:^{-}\cat C$
%% for contravariant dependence, with the opposite category being a
%% modality that internalizes flipping the variance as a type
%% constructor.
%% %
%% With these two extensions, possibly, the category contexts for
%% functors and profunctors could be unified as one notion of a sequence
%% of object variables with variances.
%% %
%% With this unification, it would then be simpler to incorporate a third
%% extension: a universe object, corresponding to an object
%% $\textrm{Set}$, which would allow us to internalize the profunctor
%% judgment in the functor judgment, modeling that in classical category
%% theory, profunctors are the same as certain functors to the category
%% of Sets.
%% %
%% While all of these extensions seem plausible, they all move away to
%% the simple ordered type theory structure that allowed us to define
%% units, tensor and hom profunctors so easily. It is unclear how to
%% define these types in this extended structure. As a simple example, if
%% profunctors have many variables of mixed variance, rather than exactly
%% one of each variance, it is unclear what the appropriate
%% generalization of the transformation context $\Phi$ should be.

%% An orthogonal extension to these would be to consider adding a
%% \emph{discretization} modality $\Delta\cat C$ that gives the discrete
%% category on the underlying set of objects of $\cat C$. This would give
%% us the ability to formalize concepts like ``function on objects'' or
%% ``unnatural transformation'' without the need to define internal
%% categories. This could be modeled by adding a modal distinction
%% $\alpha:^{\#}\cat C$ for ``crisp'' dependence on an object of $\cat
%% C$, i.e., dependence merely on the object of $\cat C$ with no
%% functoriality/naturality conditions. This would then be internalized
%% by the discretization modality.

%% A third approach would be to extend the calculus to model general
%% monoidal closed structures and more generally multicategories, which
%% provides semantics of modal substructural type theories.
%% %
%% This can be modeled by extending the ``universe'' $\Set$ to be a
%% monoidal closed category, which induces a monoidal closed structure on
%% profunctors by the Day convolution product \cite{...}.
%% %
%% Similar developments along these lines have already been done for
%% coend calculus \citep{coend-calculus} and in formalization of
%% substructural type theories \citep{atkey-woods}.

@@

%% \subsubsection*{Acknowledgements}
%% Grants stuff...? The authors would like to thank David Jaz Myers, Emily Riehl, Mike
%% Shulman, Dominic Verity for helpful feedback on this work.

@@

%% %% Please use bibtex,
